Binocular combination in the autonomic nervous system

### 2024-05-08 manuscript.Rmd
Edits to discussion

processdata <- 1  # this flag determines the amount of processing, with 4 levels:
# 0 - do no processing, generate the pdf using existing versions of all figures
# 1 - generate figures using the processed group data
# 2 - average data across participants, do bootstrapping
# 3 - download all raw data and analyse (requires 56GB of storage)

@@

condnames <- c('Luminance 1F','L-M 1F','S 1F','Melanopsin 1F','Luminance 2F','L-M 2F','S 2F','Melanopsin 2F')
zlist <- c(pupilZLum, pupilZLM, pupilZS, pupilZMel, pupilZLum2, pupilZLM2, pupilZS2, pupilZMel2)
klist <- c(pupilKLum, pupilKLM, pupilKS, pupilKMel, pupilKLum2, pupilKLM2, pupilKS2, pupilKMel2)
wlist <- c(pupilWLum, pupilWLM, pupilWS, pupilWMel, pupilWLum2, pupilWLM2, pupilWS2, pupilWMel2)
plist <- c(pupilPLum, pupilPLM, pupilPS, pupilPMel, pupilPLum2, pupilPLM2, pupilPS2, pupilPMel2)
qlist <- c(pupilQLum, pupilQLM, pupilQS, pupilQMel, pupilQLum2, pupilQLM2, pupilQS2, pupilQMel2)
Rlist <- c(pupilRmaxLum, pupilRmaxLM, pupilRmaxS, pupilRmaxMel, pupilRmaxLum2, pupilRmaxLM2, pupilRmaxS2, pupilRmaxMel2)
tabledata <- data.frame(condnames,round(zlist,digits=2),round(klist,digits=5),round(wlist,digits=2),round(plist,digits=2),round(qlist,digits=2),round(Rlist,digits=4))

@@

We used pupillometry and silent substitution to measure monocular and binocular responses of the pupils to flickering stimuli when stimulating specific photoreceptor pathways. In our luminance experiment, we found that peripheral stimulation of the retina can elicit a pupillary response and we were also able to record a response at the second harmonic, which was not seen when we stimulated the fovea [@Segala2023]. In all four experiments, we were able to record contrast response functions at both the first and the second harmonics. All experiments showed that binocular combination in the autonomic nervous system happens in a non-linear manner, with evidence of different magnitudes of interocular suppression depending on the photoreceptor pathway. This pattern of results was confirmed by a computational model, which allowed us to compare the weight of interocular suppression for each pathway. We found that, at the first harmonic frequency, the L-M pathway had a bigger suppression weight than the other pathways, while, at the second harmonic frequency, the S-(L+M) pathway had a bigger suppression weight. Finally, we looked at the pupil phases and observed different lag times for the different pathways at the first and second harmonic frequencies.

Our four experiments produced results that we were expecting based on our previous pupillometry findings [@Segala2023]: the signals from the two retinae are combined non-linearly regardless of which photoreceptor pathway is being stimulated and whether the fovea or the periphery of the retina is being stimulated. We also observed responses at both the first and the second harmonic frequencies, which were expected based on previous literature [@Spitschan2014; @Stark1957; @Verdon1988]. There was, however, one aspect of our results that we were surprised to observe: the amplitude of this second harmonic response in the different photoreceptor pathways. It is true that we had already observed second harmonic responses when stimulating the periphery of the retina with achromatic luminance modulations flickering at 2 Hz (see Figure \@ref(fig:appendixfig1)b), but the peak was very small when compared to the first harmonic peak, which is similar to what we observed in the luminance experiment here. On the other hand, in the melanopsin and L-M pathway experiments, the first and second harmonics are similar in amplitude, and in the S-(L+M) pathway experiment, the peak of the second harmonic is even bigger than the peak of the first harmonic. One reason for why this might be happening may be because the driving signal is stronger when using a lower flicker frequency [indeed, pupillometry is more responsive to slow flicker frequencies of less than 1 Hz, @Spitschan2014] and when stimulating the periphery of the retina (we have a very clear waveform for the luminance experiment in Figure \@ref(fig:pupildataLum)b). However, this explanation is not entirely satisfactory: we observed a second harmonic when stimulating the periphery using a higher flicker frequency and the pupil waveforms from the other three experiments are not as clear as the one from the luminance experiment. The presence of a second harmonic might indicate the existence of non-linearities that are not observable in the fovea: indeed, steady-state visual responses can contain activity beyond the first harmonic and these harmonics reflect non-linearities (e.g. squaring, exponentiation, rectification) in the visual response [@Norcia2015]. Additionally, it can be the case that, when these non-linearities are strong, then the higher harmonics can have equal or even higher peaks than the first harmonic. Here, we did observe strong non-linearities in our photoreceptor pathways, as reflected by the interocular suppression weights that were output by our computational model. For the S-(L+M) pathway specifically, it seems that the higher peak of the second harmonic is a behaviour intrinsic to S cones, as this was observed previously [@Spitschan2014].

The pupil phases shown in Figure \@ref(fig:phaseplots) give us more information about the pathways controlling the pupil response. The first thing that we  notice is that the luminance and L-M pathway responses are in anti-phase. It is important to note that, for our L-M pathway stimulation, the stimulus flickered between magenta (first half of the cycle) and cyan (second half of the cycle), which means that the pupil was dilating during the magenta stimulation and then constricting during the cyan presentation. This seems to suggest that increments in magenta would correspond to decrements in brightness and increments in cyan would correspond to increments in brightness. It is possible that, if we had reversed the presentation cycle for the L-M pathway, the pupil response would have been in phase with the luminance responses. It has been recently shown that L-cone and melanopsin stimulation induce an opponent pupillary response to M- and S-cone stimulation [@Murray2018; @Woelders2018]. Our melanopsin and S-(L+M) results show a similar behaviour both for monocular and binocular responses. The L-M pathways results are desynchronised approximately in quadrature phase from the melanopsin and S-(L+M) results. These results also seem to align with what was previously found: combining L and M responses should display results that are between the opponency shown by the isolated cone stimulation. Additionally, this could explain the very strong suppression that binocular (L-M) stimulation displays compared to the monocular stimulation: if L and M produce pupil responses that are in anti-phase with each other, then combining the two together should elicit smaller responses. Finally, since the results from our experiments are not in phase, it would suggest that the photoreceptors control the pupil diameter through different pathways.

@@

We have demonstrated that binocular combination of temporal flickering light in the autonomic nervous system depends on the photoreceptor pathway. We were able to elicit pupil responses by stimulating only the periphery of the retina and we were able to record contrast response functions for all photoreceptor pathways. While all pathways showed non-linear combination, they showed a variation in how the signals are combined, particularly in the weight of interocular suppression.

>>>>

processdata <- 0  # this flag determines the amount of processing, with 4 levels:
# 0 - do no processing, generate the pdf using existing versions of all figures
# 1 - generate figures using the processed group data
# 2 - average data across participants, do bootstrapping
# 3 - download all raw data and analyse (requires 56GB of storage)

@@

condnames <- c('Luminance 1F','L-M cone 1F','S cone 1F','Melanopsin 1F','Luminance 2F','L-M cone 2F','S cone 2F','Melanopsin 2F')
zlist <- c(pupilZLum, pupilZLM, pupilZS, pupilZMel, pupilZLum2, pupilZLM2, pupilZS2, pupilZMel2)
klist <- c(pupilKLum, pupilKLM, pupilKS, pupilKMel, pupilKLum2, pupilKLM2, pupilKS2, pupilKMel2)
wlist <- c(pupilWLum, pupilWLM, pupilWS, pupilWMel, pupilWLum2, pupilWLM2, pupilWS2, pupilWMel2)
plist <- c(pupilPLum, pupilPLM, pupilPS, pupilPMel, pupilPLum2, pupilPLM2, pupilPS2, pupilPMel2)
qlist <- c(pupilQLum, pupilQLM, pupilQS, pupilQMel, pupilQLum2, pupilQLM2, pupilQS2, pupilQMel2)
Rlist <- c(pupilRmaxLum, pupilRmaxLM, pupilRmaxS, pupilRmaxMel, pupilRmaxLum2, pupilRmaxLM2, pupilRmaxS2, pupilRmaxMel2)
tabledata <- data.frame(condnames,round(zlist,digits=2),round(klist,digits=5),round(wlist,digits=2),round(plist,digits=2),round(qlist,digits=2),round(Rlist,digits=4))

@@

We used binocular pupillometry and silent substitution to measure monocular and binocular responses of the pupils to flickering stimuli when stimulating specific photoreceptor pathways. In all four experiments, we were able to record contrast response functions at both the first and the second harmonic frequencies. All experiments showed that binocular combination in the autonomic nervous system happens in a non-linear manner, with evidence of different magnitudes of interocular suppression depending on the photoreceptor pathway. This pattern of results was confirmed by a computational model, which allowed us to compare the weight of interocular suppression for each pathway. We found that at the first harmonic frequency the L-M and melanopsin pathways involved strong suppression, whereas the luminance and S-cone pathways involved weaker suppression. Suppression was strong in all four pathways at the second harmonic frequency. Finally, the phase of the pupil response revealed different lag times for the different pathways at the first and second harmonic frequencies.

This is the first study to investigate binocular interactions in the melanopsin pathway directly. A previous meta-analysis [@Spitschan2019] indicated that there may be a substantial binocular facilitation effect in the circadian pathway, as indexed by melatonin suppression (melatonin is a hormone released by the pineal gland; its production is suppressed by exposure to bright light, particularly when the melanopsin-containing ipRGCs are stimulated). In brief, monocular stimulation of the ipRGCs requires up to ten times the signal strength to produce an equivalent effect to binocular stimulation. This superadditive effect implies an absence of interocular suppression, and perhaps the presence of either a highly compressive nonlinearity, or an AND-style neural operation. Our findings here are very different - we do not see substantial binocular facilitation effects in response to melanopsin modulation, and our data indicate that interocular suppression is very strong. Moreover, by measuring the full contrast response function, we can rule out compressive nonlinearities, because the functions accelerate at both the first and second harmonic frequencies. The explanation for this difference could be due to different anatomical pathways. Binocular combination in the circadian system likely takes place in the suprachiasmatic nucleus, whereas in the pupil constriction circuit the Edinger-Westphal nucleus is the most likely site of binocular integration [@Mathot2018]. Presumably these anatomical differences, and the practical constraints of the two systems, lead to differences in response.

Our recent psychophysical work [@Baker2024] has looked at binocular interactions in the L-M and S-cone pathways, and compared these to the luminance pathway. Using a contrast discrimination paradigm with spatial modulations of luminance and colour, @Baker2024 found equally strong interocular suppression in all three pathways. This is rather different from our pupillometry results here, which demonstrate weaker suppression in the luminance and S-cone pathways than in the L-M pathway (see Figure \@ref(fig:FirstHarmonicPlots)). However, the current experiments involve temporal modulations, which are quite distinct from the spatial modulations used in @Baker2024. Our other recent work [@Segala2023] has shown that temporal luminance modulations involve much weaker interocular suppression in the cortical response than do spatial luminance modulations [@Baker2017]. These different normalization processes might reflect different priorities for spatial and temporal vision. Spatial vision aims to fuse images to provide binocular single vision, and benefits from 'ocularity invariance' [@Baker2007], in which visual appearance is constant when viewed with one eye or two. Temporal vision is critical for motion perception, which can involve alternative binocular computations, such as calculating velocity differences between the eyes [@Kaestner2019]. Of course the present measurements were of pupil size, which may be subject to different anatomical and functional constraints from those in the cortex. Future research should aim to extend our current findings to perception using psychophysical approaches.

@@

We have demonstrated that binocular combination of temporal flickering light in the autonomic nervous system depends on the photoreceptor pathway. We were able to elicit pupil responses by stimulating the periphery of the retina and we were able to record contrast response functions for all photoreceptor pathways. While all pathways showed non-linear combination, they varied in how the signals are combined, particularly in the weight of interocular suppression. This was strong ($\omega \ge 1$) for L-M and melanopsin signals at the first harmonic, and all pathways at the second harmonic. Suppression was weaker ($\omega < 1$) in the luminance pathway (consistent with previous work), and also for S-cone directed modulations.

### 2024-05-08 manuscript.tex
Edits to discussion

Different classes of retinal photoreceptors, including cones, rods and melanopsin-containing intrinsically photosensitive retinal ganglion cells (ipRGCs), are directly involved in controlling and maintaining the size of the pupils (\protect\hyperlink{ref-Barrionuevo2018}{Barrionuevo et al., 2018}; \protect\hyperlink{ref-Barrionuevo2014}{Barrionuevo et al., 2014}; \protect\hyperlink{ref-Dacey2005}{Dacey et al., 2005}; \protect\hyperlink{ref-Murray2018}{Murray et al., 2018}; \protect\hyperlink{ref-Spitschan2019}{Spitschan, 2019}; \protect\hyperlink{ref-Spitschan2014}{Spitschan et al., 2014}; \protect\hyperlink{ref-Woelders2018}{Woelders et al., 2018}). The cones have been shown to drive the initial rapid constriction of the pupils (\protect\hyperlink{ref-Mathot2018}{Mathôt, 2018}), while the slower and longer activation of the ipRGCs shows their role in maintaining the constriction over a prolonged period of time and in regulating the post-illumination pupillary response (\protect\hyperlink{ref-Markwell2010}{Markwell et al., 2010}; \protect\hyperlink{ref-McDougal2010}{McDougal and Gamlin, 2010}). The ipRGCs are a unique photoreceptor class that has been discovered recently (\protect\hyperlink{ref-Provencio2000}{Provencio et al., 2000}). They express the photopigment melanopsin, and are involved in the regulation of the circadian rhythm (\protect\hyperlink{ref-Panda2002}{Panda et al., 2002}; \protect\hyperlink{ref-Provencio2000}{Provencio et al., 2000}; \protect\hyperlink{ref-Ruby2002}{Ruby et al., 2002}), forming a major input to the PON (\protect\hyperlink{ref-Dacey2003}{Dacey et al., 2003}). The first direct evidence of the involvement of the ipRGCs in the PLR was shown in rats by Lucas et al. (\protect\hyperlink{ref-Lucas2003}{2003}) who genetically deleted the melanopsin gene from rats, resulting in the loss of the intrinsic photosensitivity of the cells and a reduced pupil constriction. This same behaviour was later also observed in primates and humans (\protect\hyperlink{ref-Gamlin2007}{Gamlin et al., 2007}), where it was demonstrated that the PLR continues during light presentation even when cone and rod signalling is blocked, indicating the primary role of the ipRGCs in maintaining pupil constriction over a prolonged time.

@@

\caption{\label{tab:paramtable}Summary of maximum a posteriori (MAP) parameter estimates for each data set.}
\centering
\begin{tabular}[t]{lcccccc}
\toprule
Experiment & Z & k & w & p & q & Rmax\\
\midrule
Luminance 1F & 95.77 & 0.00083 & 0.37 & 1.32 & 1.24 & 0.0905\\
L-M 1F & 49.82 & 0.00109 & 1.71 & 1.50 & 1.21 & 0.0032\\
S 1F & 63.00 & 0.00121 & 0.39 & 1.94 & 1.79 & 0.0042\\
Melanopsin 1F & 79.61 & 0.00094 & 1.71 & 1.29 & 0.76 & 0.0025\\
\midrule
Luminance 2F & 48.81 & 0.00089 & 1.24 & 1.15 & 0.88 & 0.0039\\
L-M 2F & 66.00 & 0.00075 & 0.95 & 1.10 & 0.82 & 0.0050\\
S 2F & 75.94 & 0.00097 & 2.22 & 1.41 & 0.83 & 0.0018\\
Melanopsin 2F & 67.94 & 0.00101 & 1.37 & 0.90 & 0.49 & 0.0035\\
\bottomrule
\end{tabular}
\end{table}

@@

We used pupillometry and silent substitution to measure monocular and binocular responses of the pupils to flickering stimuli when stimulating specific photoreceptor pathways. In our luminance experiment, we found that peripheral stimulation of the retina can elicit a pupillary response and we were also able to record a response at the second harmonic, which was not seen when we stimulated the fovea (\protect\hyperlink{ref-Segala2023}{Segala et al., 2023}). In all four experiments, we were able to record contrast response functions at both the first and the second harmonics. All experiments showed that binocular combination in the autonomic nervous system happens in a non-linear manner, with evidence of different magnitudes of interocular suppression depending on the photoreceptor pathway. This pattern of results was confirmed by a computational model, which allowed us to compare the weight of interocular suppression for each pathway. We found that, at the first harmonic frequency, the L-M pathway had a bigger suppression weight than the other pathways, while, at the second harmonic frequency, the S-(L+M) pathway had a bigger suppression weight. Finally, we looked at the pupil phases and observed different lag times for the different pathways at the first and second harmonic frequencies.

Our four experiments produced results that we were expecting based on our previous pupillometry findings (\protect\hyperlink{ref-Segala2023}{Segala et al., 2023}): the signals from the two retinae are combined non-linearly regardless of which photoreceptor pathway is being stimulated and whether the fovea or the periphery of the retina is being stimulated. We also observed responses at both the first and the second harmonic frequencies, which were expected based on previous literature (\protect\hyperlink{ref-Spitschan2014}{Spitschan et al., 2014}; \protect\hyperlink{ref-Stark1957}{Stark and Sherman, 1957}; \protect\hyperlink{ref-Verdon1988}{Verdon and Howarth, 1988}). There was, however, one aspect of our results that we were surprised to observe: the amplitude of this second harmonic response in the different photoreceptor pathways. It is true that we had already observed second harmonic responses when stimulating the periphery of the retina with achromatic luminance modulations flickering at 2 Hz (see Figure \ref{fig:appendixfig1}b), but the peak was very small when compared to the first harmonic peak, which is similar to what we observed in the luminance experiment here. On the other hand, in the melanopsin and L-M pathway experiments, the first and second harmonics are similar in amplitude, and in the S-(L+M) pathway experiment, the peak of the second harmonic is even bigger than the peak of the first harmonic. One reason for why this might be happening may be because the driving signal is stronger when using a lower flicker frequency (indeed, pupillometry is more responsive to slow flicker frequencies of less than 1 Hz, \protect\hyperlink{ref-Spitschan2014}{Spitschan et al., 2014}) and when stimulating the periphery of the retina (we have a very clear waveform for the luminance experiment in Figure \ref{fig:pupildataLum}b). However, this explanation is not entirely satisfactory: we observed a second harmonic when stimulating the periphery using a higher flicker frequency and the pupil waveforms from the other three experiments are not as clear as the one from the luminance experiment. The presence of a second harmonic might indicate the existence of non-linearities that are not observable in the fovea: indeed, steady-state visual responses can contain activity beyond the first harmonic and these harmonics reflect non-linearities (e.g.~squaring, exponentiation, rectification) in the visual response (\protect\hyperlink{ref-Norcia2015}{Norcia et al., 2015}). Additionally, it can be the case that, when these non-linearities are strong, then the higher harmonics can have equal or even higher peaks than the first harmonic. Here, we did observe strong non-linearities in our photoreceptor pathways, as reflected by the interocular suppression weights that were output by our computational model. For the S-(L+M) pathway specifically, it seems that the higher peak of the second harmonic is a behaviour intrinsic to S cones, as this was observed previously (\protect\hyperlink{ref-Spitschan2014}{Spitschan et al., 2014}).

The pupil phases shown in Figure \ref{fig:phaseplots} give us more information about the pathways controlling the pupil response. The first thing that we notice is that the luminance and L-M pathway responses are in anti-phase. It is important to note that, for our L-M pathway stimulation, the stimulus flickered between magenta (first half of the cycle) and cyan (second half of the cycle), which means that the pupil was dilating during the magenta stimulation and then constricting during the cyan presentation. This seems to suggest that increments in magenta would correspond to decrements in brightness and increments in cyan would correspond to increments in brightness. It is possible that, if we had reversed the presentation cycle for the L-M pathway, the pupil response would have been in phase with the luminance responses. It has been recently shown that L-cone and melanopsin stimulation induce an opponent pupillary response to M- and S-cone stimulation (\protect\hyperlink{ref-Murray2018}{Murray et al., 2018}; \protect\hyperlink{ref-Woelders2018}{Woelders et al., 2018}). Our melanopsin and S-(L+M) results show a similar behaviour both for monocular and binocular responses. The L-M pathways results are desynchronised approximately in quadrature phase from the melanopsin and S-(L+M) results. These results also seem to align with what was previously found: combining L and M responses should display results that are between the opponency shown by the isolated cone stimulation. Additionally, this could explain the very strong suppression that binocular (L-M) stimulation displays compared to the monocular stimulation: if L and M produce pupil responses that are in anti-phase with each other, then combining the two together should elicit smaller responses. Finally, since the results from our experiments are not in phase, it would suggest that the photoreceptors control the pupil diameter through different pathways.

@@

We have demonstrated that binocular combination of temporal flickering light in the autonomic nervous system depends on the photoreceptor pathway. We were able to elicit pupil responses by stimulating only the periphery of the retina and we were able to record contrast response functions for all photoreceptor pathways. While all pathways showed non-linear combination, they showed a variation in how the signals are combined, particularly in the weight of interocular suppression.

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-Barrionuevo2018}{}}%
Barrionuevo PA, McAnany JJ, Zele AJ, Cao D. 2018. \href{https://www.frontiersin.org/articles/10.3389/fneur.2018.01140}{Non-linearities in the {Rod} and {Cone} {Photoreceptor} {Inputs} to the {Afferent} {Pupil} {Light} {Response}}. \emph{Frontiers in Neurology} \textbf{9}.

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-Spitschan2019}{}}%
Spitschan M. 2019. Photoreceptor inputs to pupil control. \emph{Journal of Vision} \textbf{19}:5. doi:\href{https://doi.org/10.1167/19.9.5}{10.1167/19.9.5}

@@

\leavevmode\vadjust pre{\hypertarget{ref-Spitschan2018}{}}%
Spitschan M, Woelders T. 2018. \href{https://www.frontiersin.org/articles/10.3389/fneur.2018.00941}{The {Method} of {Silent} {Substitution} for {Examining} {Melanopsin} {Contributions} to {Pupil} {Control}}. \emph{Frontiers in Neurology} \textbf{9}.

\leavevmode\vadjust pre{\hypertarget{ref-Stark1957}{}}%
Stark L, Sherman PM. 1957. A servoanalytic study of consensual pupil reflex to light. \emph{Journal of Neurophysiology} \textbf{20}:17--26. doi:\href{https://doi.org/10.1152/jn.1957.20.1.17}{10.1152/jn.1957.20.1.17}

@@

\leavevmode\vadjust pre{\hypertarget{ref-Verdon1988}{}}%
Verdon W, Howarth PA. 1988. The pupil's response to short wavelength cone stimulation. \emph{Vision research} \textbf{28}:1119--1128. doi:\href{https://doi.org/10.1016/0042-6989(88)90138-1}{10.1016/0042-6989(88)90138-1}

>>>>

Different classes of retinal photoreceptors, including cones, rods and melanopsin-containing intrinsically photosensitive retinal ganglion cells (ipRGCs), are directly involved in controlling and maintaining the size of the pupils (\protect\hyperlink{ref-Barrionuevo2018}{Barrionuevo et al., 2018}; \protect\hyperlink{ref-Barrionuevo2014}{Barrionuevo et al., 2014}; \protect\hyperlink{ref-Dacey2005}{Dacey et al., 2005}; \protect\hyperlink{ref-Murray2018}{Murray et al., 2018}; \protect\hyperlink{ref-Spitschan2014}{Spitschan et al., 2014}; \protect\hyperlink{ref-Spitschan2019}{Spitschan and Cajochen, 2019}; \protect\hyperlink{ref-Woelders2018}{Woelders et al., 2018}). The cones have been shown to drive the initial rapid constriction of the pupils (\protect\hyperlink{ref-Mathot2018}{Mathôt, 2018}), while the slower and longer activation of the ipRGCs shows their role in maintaining the constriction over a prolonged period of time and in regulating the post-illumination pupillary response (\protect\hyperlink{ref-Markwell2010}{Markwell et al., 2010}; \protect\hyperlink{ref-McDougal2010}{McDougal and Gamlin, 2010}). The ipRGCs are a unique photoreceptor class that has been discovered recently (\protect\hyperlink{ref-Provencio2000}{Provencio et al., 2000}). They express the photopigment melanopsin, and are involved in the regulation of the circadian rhythm (\protect\hyperlink{ref-Panda2002}{Panda et al., 2002}; \protect\hyperlink{ref-Provencio2000}{Provencio et al., 2000}; \protect\hyperlink{ref-Ruby2002}{Ruby et al., 2002}), forming a major input to the PON (\protect\hyperlink{ref-Dacey2003}{Dacey et al., 2003}). The first direct evidence of the involvement of the ipRGCs in the PLR was shown in rats by Lucas et al. (\protect\hyperlink{ref-Lucas2003}{2003}) who genetically deleted the melanopsin gene from rats, resulting in the loss of the intrinsic photosensitivity of the cells and a reduced pupil constriction. This same behaviour was later also observed in primates and humans (\protect\hyperlink{ref-Gamlin2007}{Gamlin et al., 2007}), where it was demonstrated that the PLR continues during light presentation even when cone and rod signalling is blocked, indicating the primary role of the ipRGCs in maintaining pupil constriction over a prolonged time.

@@

\caption{\label{tab:paramtable}Summary of maximum a posteriori (MAP) parameter estimates for each data set.}
\centering
\begin{tabular}[t]{lcccccc}
\toprule
Experiment & Z & k & w & p & q & Rmax\\
\midrule
Luminance 1F & 95.77 & 0.00083 & 0.37 & 1.32 & 1.24 & 0.0905\\
L-M cone 1F & 49.82 & 0.00109 & 1.71 & 1.50 & 1.21 & 0.0032\\
S cone 1F & 63.00 & 0.00121 & 0.39 & 1.94 & 1.79 & 0.0042\\
Melanopsin 1F & 79.61 & 0.00094 & 1.71 & 1.29 & 0.76 & 0.0025\\
\midrule
Luminance 2F & 48.81 & 0.00089 & 1.24 & 1.15 & 0.88 & 0.0039\\
L-M cone 2F & 66.00 & 0.00075 & 0.95 & 1.10 & 0.82 & 0.0050\\
S cone 2F & 75.94 & 0.00097 & 2.22 & 1.41 & 0.83 & 0.0018\\
Melanopsin 2F & 67.94 & 0.00101 & 1.37 & 0.90 & 0.49 & 0.0035\\
\bottomrule
\end{tabular}
\end{table}

@@

We used binocular pupillometry and silent substitution to measure monocular and binocular responses of the pupils to flickering stimuli when stimulating specific photoreceptor pathways. In all four experiments, we were able to record contrast response functions at both the first and the second harmonic frequencies. All experiments showed that binocular combination in the autonomic nervous system happens in a non-linear manner, with evidence of different magnitudes of interocular suppression depending on the photoreceptor pathway. This pattern of results was confirmed by a computational model, which allowed us to compare the weight of interocular suppression for each pathway. We found that at the first harmonic frequency the L-M and melanopsin pathways involved strong suppression, whereas the luminance and S-cone pathways involved weaker suppression. Suppression was strong in all four pathways at the second harmonic frequency. Finally, the phase of the pupil response revealed different lag times for the different pathways at the first and second harmonic frequencies.

This is the first study to investigate binocular interactions in the melanopsin pathway directly. A previous meta-analysis (\protect\hyperlink{ref-Spitschan2019}{Spitschan and Cajochen, 2019}) indicated that there may be a substantial binocular facilitation effect in the circadian pathway, as indexed by melatonin suppression (melatonin is a hormone released by the pineal gland; its production is suppressed by exposure to bright light, particularly when the melanopsin-containing ipRGCs are stimulated). In brief, monocular stimulation of the ipRGCs requires up to ten times the signal strength to produce an equivalent effect to binocular stimulation. This superadditive effect implies an absence of interocular suppression, and perhaps the presence of either a highly compressive nonlinearity, or an AND-style neural operation. Our findings here are very different - we do not see substantial binocular facilitation effects in response to melanopsin modulation, and our data indicate that interocular suppression is very strong. Moreover, by measuring the full contrast response function, we can rule out compressive nonlinearities, because the functions accelerate at both the first and second harmonic frequencies. The explanation for this difference could be due to different anatomical pathways. Binocular combination in the circadian system likely takes place in the suprachiasmatic nucleus, whereas in the pupil constriction circuit the Edinger-Westphal nucleus is the most likely site of binocular integration (\protect\hyperlink{ref-Mathot2018}{Mathôt, 2018}). Presumably these anatomical differences, and the practical constraints of the two systems, lead to differences in response.

Our recent psychophysical work (\protect\hyperlink{ref-Baker2024}{Baker et al., 2024}) has looked at binocular interactions in the L-M and S-cone pathways, and compared these to the luminance pathway. Using a contrast discrimination paradigm with spatial modulations of luminance and colour, Baker et al. (\protect\hyperlink{ref-Baker2024}{2024}) found equally strong interocular suppression in all three pathways. This is rather different from our pupillometry results here, which demonstrate weaker suppression in the luminance and S-cone pathways than in the L-M pathway (see Figure \ref{fig:FirstHarmonicPlots}). However, the current experiments involve temporal modulations, which are quite distinct from the spatial modulations used in Baker et al. (\protect\hyperlink{ref-Baker2024}{2024}). Our other recent work (\protect\hyperlink{ref-Segala2023}{Segala et al., 2023}) has shown that temporal luminance modulations involve much weaker interocular suppression in the cortical response than do spatial luminance modulations (\protect\hyperlink{ref-Baker2017}{Baker and Wade, 2017}). These different normalization processes might reflect different priorities for spatial and temporal vision. Spatial vision aims to fuse images to provide binocular single vision, and benefits from `ocularity invariance' (\protect\hyperlink{ref-Baker2007}{Baker et al., 2007}), in which visual appearance is constant when viewed with one eye or two. Temporal vision is critical for motion perception, which can involve alternative binocular computations, such as calculating velocity differences between the eyes (\protect\hyperlink{ref-Kaestner2019}{Kaestner et al., 2019}). Of course the present measurements were of pupil size, which may be subject to different anatomical and functional constraints from those in the cortex. Future research should aim to extend our current findings to perception using psychophysical approaches.

@@

We have demonstrated that binocular combination of temporal flickering light in the autonomic nervous system depends on the photoreceptor pathway. We were able to elicit pupil responses by stimulating the periphery of the retina and we were able to record contrast response functions for all photoreceptor pathways. While all pathways showed non-linear combination, they varied in how the signals are combined, particularly in the weight of interocular suppression. This was strong (\(\omega \ge 1\)) for L-M and melanopsin signals at the first harmonic, and all pathways at the second harmonic. Suppression was weaker (\(\omega < 1\)) in the luminance pathway (consistent with previous work), and also for S-cone directed modulations.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Baker2024}{}}%
Baker DH, Hansford KJ, Segala FG, Morsi AY, Huxley RJ, Martin JT, Rockman M, Wade AR. 2024. Binocular integration of chromatic and luminance signals. \emph{Journal of Vision} \textbf{Under review}. doi:\href{https://doi.org/10.31234/osf.io/atsfh}{10.31234/osf.io/atsfh}

@@

\leavevmode\vadjust pre{\hypertarget{ref-Barrionuevo2018}{}}%
Barrionuevo PA, McAnany JJ, Zele AJ, Cao D. 2018. Non-linearities in the {Rod} and {Cone} {Photoreceptor} {Inputs} to the {Afferent} {Pupil} {Light} {Response}. \emph{Frontiers in Neurology} \textbf{9}. doi:\href{https://doi.org/10.3389/fneur.2018.01140}{10.3389/fneur.2018.01140}

@@

\leavevmode\vadjust pre{\hypertarget{ref-Kaestner2019}{}}%
Kaestner M, Maloney RT, Wailes-Newson KH, Bloj M, Harris JM, Morland AB, Wade AR. 2019. Asymmetries between achromatic and chromatic extraction of 3D motion signals. \emph{Proc Natl Acad Sci U S A} \textbf{116}:13631--13640. doi:\href{https://doi.org/10.1073/pnas.1817202116}{10.1073/pnas.1817202116}

@@

\leavevmode\vadjust pre{\hypertarget{ref-Spitschan2019}{}}%
Spitschan M, Cajochen C. 2019. Binocular facilitation in light-mediated melatonin suppression? \emph{J Pineal Res} \textbf{67}:e12602. doi:\href{https://doi.org/10.1111/jpi.12602}{10.1111/jpi.12602}

@@

\leavevmode\vadjust pre{\hypertarget{ref-Spitschan2018}{}}%
Spitschan M, Woelders T. 2018. The {Method} of {Silent} {Substitution} for {Examining} {Melanopsin} {Contributions} to {Pupil} {Control}. \emph{Frontiers in Neurology} \textbf{9}. doi:\href{https://doi.org/10.3389/fneur.2018.00941}{10.3389/fneur.2018.00941}

@@
